---
title: "Introduction to RTMB"
author: "Kasper Kristensen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE}
##library(knitr)
library(RTMB)
set.seed(1)
formals(MakeADFun)$silent <- TRUE
```

## Content

- type invariant implementation
- parameter branching
- methods that can be used
- simulation: catching common errors
- TMB interface: ADREPORT, REPORT,...
- OSA-residuals
- idea model: ar1xar1 plus noise fit to volcano data
- density namespace


## An example model

```{r ar1xar1}
data(volcano)
data <- list(y = volcano)
parameters <- list(u=array(0, dim(volcano)), phi1=c(.1), phi2=c(.1), mu=0, sd.field=1, sd.meas=1)
func <- function(parameters) {
    getAll(parameters, data)
    y <- OBS(y)
    ## Autoregresssive models
    f1 <- function(x)dautoreg(x, phi=phi1, log=TRUE)
    f2 <- function(x)dautoreg(x, phi=phi2, log=TRUE)
    jnll <- -dseparable(f1, f2)(u)
    jnll <- jnll - sum(dnorm(y, mu + sd.field*u, sd=sd.meas, log=TRUE))
    jnll
}
```

First, we check that we can evaluate the function for standard numeric inputs:

```{r}
func(parameters)
```

## The TMB interface

Next, we setup the TMB objective using `MakeADFun()` and fit the model using the recommended optimizer `nlminb`:

```{r}
obj <- MakeADFun(func, parameters, random="u")
opt <- nlminb(obj$par, obj$fn, obj$gr)
```

During the optimization it is common that the optimizer `nlminb` evaluates the objective function for parameters that are extreme. This may cause warnings such as

> Warning messages:
> 1: In nlminb(obj$par, obj$fn, obj$gr) : NA/NaN function evaluation

which is harmless as long as there are only a few of them.

The TMB function `sdreport` gives a standard summary of the fitted object:

```{r}
sdr <- sdreport(obj)
sdr
```

## Simulating from the model object


```{r}
sim <- obj$simulate()

```

What if we want to try AR2 instead of AR1? Set


```{r}
parameters$phi1 <- c(.1,0)
parameters$phi2 <- c(.1,0)

```

and repeat the analysis.
